Quantitative Benchmarking for Automatically Generated Conjectures

### 2017-11-13 TEBenchmarkPaper/article.tex
Add timing graph
This is from example runs at the moment (few repetitions, short cutoffs).

The scripts should apply generally though.
Much addition

The precision and recall for multiple samples $S$ can be combined in two ways:
finding their mean value (the \emph{average of the ratios}) or by summing the
output and ground-truth counts \emph{before} taking their ratios (the
\emph{ratio of the averages}). The former estimates the precision and recall we
can expect from any \emph{particular} run, which is of direct relevance to users
of these tools. The latter estimates the proportion of discoveries that a tool
can expect to find out of 'all possible', i.e. assigning higher weight to
performance in theories which have more conjectures available to be discovered;
whilst this is of theoretical interest, it is a more difficult claim to support
empirically, so we leave its investigation to future work.

@@

% Better presentation, flowchart?
\begin{itemize}
\item Collect together all definitions found in TIP into one large theory,
  removing $\alpha$-equivalent duplicates.
\item For each type, provide extra functions which act as constructors and
  destructors. For example, the $\List$ type would have extra constructor
  functions $\text{constructor-Nil}$ and $\text{constructor-Cons}$, and
  destructor functions $\text{destructor-head}$ and $\text{destructor-tail}$.
  This allows us to ignore constructors and destructors, and concentrate solely
  on functions.
\item Collect together all theorem statements found in TIP. Update any
  references to removed definitions, and replace references to constructors with
  the associated constructor function from above.
\item \emph{Sample} a sub-set of functions from the large theory uniformly, such
  that we choose all of the names referenced by at least one of the theorems
  (for example, theorem \ref{eq:mapreduce} references the names
  $\Map$ and $\Append$).
\item Provide this sub-set as an input to the ATE system, timing how long it
  takes to complete.
\item Perform precision/recall analysis on the resulting conjectures, against
  the ground truth of TIP theorems which only reference names in this sample.
\end{itemize}

>>>>

The precision and recall for multiple samples $S$ can be combined in two ways:
finding their mean value (the \emph{average of the ratios}) or by summing the
numerators and denominators (the \emph{ratio of the averages}). The former gives
the expected precision and recall for a run, which is of direct relevance to
users of these tools, and is hence our preferred measure. The latter assigns more weight to theories which generate more
conjectures, but that is not justification there is no reason to consider these theories ``better'',

@@

We have applied our methodology to evaluating QuickSpec, %TODO:cite
a library for conjecturing equations about functions in the Haskell programming
language. We use the Tons of Inductive Problems (TIP) theorem proving benchmark
to generate our corpus.

\subsection{QuickSpec}

We use QuickSpec version 0.9.6 to demonstrate our benchmarking methodology. This
is a library for the Haskell programming language, which is given a
\emph{signature} containing Haskell functions and (universally quantified) typed
variables, and outputs conjectured equations involving these terms.

The QuickSpec algorithm first enumerates all well-typed expressions, up to a
certain depth (3 by default), involving the given terms. Expressions of the same
type are assumed to be equivalent, and grouped into equivalence classes. A
testing process then tries to disprove this assumption by instantiating each
variable with a randomly chosen value (using the QuickCheck testing library),
evaluating the expressions and comparing the results. The equivalence classes
are split to separate any expressions which are observed to be different, and
the process repeats with new random values.

After 500 rounds of testing, any expressions still sharing the same equivalence
class are conjectured to be equal. Finally, a congruence closure algorithm is
applied to these equations, to find a minimal set.

In order to thoroughly benchmark QuickSpec, we need to automate some of the
decisions which are normally left up to the user:

\begin{itemize}
\item We must decide what variables to include. We add three for each type that
  appears as a function argument, except for types which have no QuickCheck data
  generator.
\item We must \emph{monomorphise} all types. For example, functions like
  \texttt{cons} are \emph{polymorphic}: they build lists of any element type.
  We need to pick a specific type in order to generate random values, so we
  resolve this, arbitrarily, by picking
  \emph{Integer}.~\footnote{We pick \texttt{Integer} for variables of kind
    \texttt{*}, and \texttt{[]} (Haskell's list type constructor) for those of
    kind \texttt{* -> *}. If these wouldn't satisfy some constraint, we pick a
    suitable type non-deterministically from the module scope, or skip these
    variables completely if none is found.}
\item Haskell functions only take one argument, as as in the $\lambda$-calculus,
  so functions of multiple arguments must be simulated by currying.
  Unfortunately QuickSpec cannot compare functions for equality, so we must give
  each function in our signature an explicit arity. Arities from zero to five
  are supported, so we avoid partial application by picking the highest that is
  type-correct.
\end{itemize}

To find out which functions to include in a signature, we use a plugin for the
GHC compiler which tells us all of the functions defined in a package. We
include all of those which type-check when passed to QuickSpec.

\subsection{TIP}

We use TIP version 0.2 which contains 343 problems, each stating a single
theorem and together defining a total of 618 datatypes and 1498 functions. Most
of these are duplicates, since each problem (re-)defines all of the datatypes
and functions it involves. We prefix the names of all definitions with the
filename it came from, and collect them into one big set.

Datatypes can have several ``constructors'' and ``destructors''. For example the
natural numbers can be defined as follows:

\begin{verbatim}
(declare-datatypes () ((Nat (Z) (S (p Nat)))))
\end{verbatim}

This defines the type \texttt{Nat}, along with the constructors \texttt{Z} (for
zero) and \texttt{S} (the successor function), and destructor \texttt{p} (the
predecessor function).

Since our GHC plugin only lists function definitions, we add separate function
definitions for each constructor and destructor (these are just
$\eta$-expansions).

TIP has ``native'' support for booleans and integers, allowing functions like
\texttt{+} to be used without an accompanying definition. Without a definition
these functions will not be included in the QuickSpec exploration, so we replace
all occurrences with custom definitions.

This gives us 3598 definitions in total, with 269 remaining after
$\alpha$-equivalent duplicates are removed.

TIP includes a tool to translate its definitions into Haskell code, including
suitable comparison functions and random data generators. We translated all 269
definitions into a single Haskell package, compiled it with GHC and extracted
the function definitions with our plugin. Each benchmark run imported this
entire package, but the signature given to QuickSpec only included functions
chosen for the current sample.

\subsection{Results}

\begin{figure}
  \includegraphics[width=8cm]{graphs/time}
  \includegraphics[width=8cm]{graphs/precRec}
  \caption{}
  \label{figure:quickspec_results}
\end{figure}

We applied our sampling methodology to the 269 TIP definitions for sample sizes
between 1 and 20, above which point QuickSpec becomes very slow. We chose 30
samples for each size, but the smaller sizes included repetitions due to the
requirement for samples to admit at least one theorem. The results are shown in
figure~\ref{figure:quickspec_results}.

Precision and recall were calculated by normalising each equation generated by
QuickSpec: all variables are identified numerically, in ascending order from
left to right; equations were also sorted such that the expression on the
left-hand-side is lexicographically smaller than that on the right.

Theorem statements from TIP were first updated to reference only the 269
distinct names we explored, and references to constructors and destructors were
updated to use the $\eta$-expanded versions instead. Equations were normalised
in the same way as above (non-equations were kept unchanged), and elements of
these two sets were deemed to match if they were syntactically identical.
